---
title: "COVID19_3"
author: "Meltem Ozcan"
date: "6/2/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(mcmc);library(dplyr);library(RColorBrewer);library(LearnBayes)
library(kableExtra);library(MASS);library(monomvn); library(lars); 
library(glmnet); library(miscTools) 
#library(plotrix)  # for drawing circles
```

```{r}
set.seed(87)
covid = read.delim('Covid19.txt', sep=',')
covid %>% glimpse()
#table(covid$Group)
#table(covid$Region)
```

1. Consider the ten census regions and group the counties into five regions using the following grouping: Group 1, counties in regions 1 and 2; Group 2, counties in region 3; Group 3, counties in region 5; Group 4, counties in regions 4 and 6; Group 5, counties in regions 7,8,9, and 10.

2. Let $y_{ij} = log n_{ij}$ where $n_{ij}$ is the number of cases, and let $c_{ij}$ the population, for the i-th county in the j-th region. Perform an exploratory data analysis involving $y_{ij}$ , the five different groups, the population, and the population density. Discuss possible associations and clusters.
```{r}
covid["lncases"]=log(covid$Total.cases)
covid["lpop"]=log(covid$Population)
covid["ldensity"]=log(covid$Density)

covid["abbr"]=c("LA","SD", "RIV","SCL" , "ORA","SBD","SF","ALA" ,"SAC","SM" , "COS", "KER",  
"VEN", "SJ","SB", "TUL", "FRE", "MRN" ,"SON" ,  "STA",
"SOL" , "PLA" ,"SLO" ,"YOL" ,   "IMP" , "SCR" ,"MON", "MER" , 
"HUM" , "SBT" ,  "ED","NAP","NEV", "MAD" ,"SHA" ,  "MON",       
"SUT" , "YUB"  , "BUT" ,"INY","KIN"  ,"CAL",   "AMA" , "SIS",   
"GLE"  ,"LK" , "MEN"  ,"PLU","COL"  , "DN", "ALP" , "TEH",     
"TUO" , "LAS" , "MPA" , "MOD","SIE"  , "TRI")

covid[covid$lncases=="-Inf",]$lncases=0

cols=c("red2","goldenrod2","palegreen3","cornflowerblue","darkorchid1")
myColors=rep(NA, 58)
myColors[which(covid$Group=="1")]="red2"
myColors[which(covid$Group=="2")]="goldenrod2"
myColors[which(covid$Group=="3")]="palegreen3"
myColors[which(covid$Group=="4")]="cornflowerblue"
myColors[which(covid$Group=="5")]="darkorchid1"
```

#EDA PLOTS
```{r}
par(las=2)
plot(covid$lncases, xlab="County", cex=0.6, bty="n",col=myColors, pch=16, 
     ylab="Log total cases",main="Log total cases by group and county", xaxt="n")
legend("topright",fill=cols, legend=c("Group 1","Group 2", "Group 3", "Group 4", "Group 5"))
axis(1, at=1:58, labels=covid$abbr,cex.axis=0.5, font=2)

par(mfrow=c(1,2),las=2)
plot(x=covid$ldensity,y=covid$lncases, xlab="Log density",
     ylab="Log total cases", col=myColors, pch=16,  main="Log n_ij vs log d_ij by group")
legend("topleft",fill=cols, legend=c("Group 1","Group 2", "Group 3", "Group 4", "Group 5"), cex=0.6, bty="n")

#G1=tapply(covid$lpop, covid$Group, mean)
plot(x=covid$lpop,y=covid$lncases, xlab="Log population",
     ylab="Log total cases", col=myColors,pch=16,  main="Log n_ij vs log c_ij by group")
legend("topleft",fill=cols, legend=c("Group 1","Group 2", "Group 3", "Group 4", "Group 5"), cex=0.6, bty="n")

#for(i in 1:length(G1)){draw.circle(G1[i], G2[i], 1, border=cols[i],lty=2,lwd=0.9)}
```

```{r}
#some table stuff
a=covid%>% group_by(Group)%>%summarise_at(vars(lncases),funs(mean,sd))
b2=covid%>%group_by(Group)%>%summarise_at(vars(ldensity),funs(mean,sd))

t1=as.data.frame(cbind(round(a[2:3],2),round(b2[2:3],2)))
t2=t(t1) %>% kable(col.names = c("Group 1", "Group 2","Group 3", "Group 4", "Group 5"))%>%
                kable_styling(bootstrap_options = c("striped"), latex_options = "hold_position")
```


############################################################################
MODELS

3. Consider, fit and assess the following models, using a Bayesian approach with priors based on your best judgement:
$$y_{ij} = \mu + \epsilon_{ij}$$
$$y_{ij} = \mu + \beta d_{ij} +  \epsilon_{ij}$$
where $d_{ij}$ is the population density of the ith county in the jth group;
$$y_{ij} = \mu +\eta_j + \epsilon_{ij}$$
where $\eta_j$ is a group effect
$$y_{ij} = \mu + \eta_j + dij +  \epsilon_{ij}$$
In all cases
$$\epsilon_{ij} \sim N(0, \sigma^2\sqrt{10^3/c_{ij}})$$


MODEL 1:
```{r}
V1=diag(1000/covid$Population) 
Linv=diag(sqrt(covid$Population/1000))

X1 = matrix(rep(1,58))
n1=nrow(X1)
k1=ncol(X1)
y1=matrix(covid$lncases,n1,1)

z1=Linv%*%y1
w1=Linv%*%X1
#qr decomposition
qr1=qr(w1)

#check for posterior propriety
stopifnot(n1>k1, qr1$rank==k1)

#calculate posterior hyperparameters
Rinv1=solve(qr.R(qr1))#qr.r extracts the upper triangular matrix
vbeta1=Rinv1%*%t(Rinv1)
betahat1=qr.solve(qr1,z1)

df1=n1-k1
e1=qr.resid(qr1,z1)
sig1=sum(e1^2)/df1

#simulate from posterior
niter=1000
sigma1=sqrt(1/rgamma(niter,df1/2,df1*sig1/2))
beta1=matrix(betahat1,niter,k1, byrow=T)+sigma1*mvrnorm(niter, rep(0,k1),vbeta1)


colMeans(beta1)
betahat1
```


```{r}
#plot beta estimates
par(las=2)
par(mfrow=c(1,2))
hist(sigma1,100, freq=F, main="Standard deviation", xlab=expression(sigma), col="lightblue", lty="blank")
hist(beta1[,1],100, freq=F, main="Intercept", xlab=expression(mu), col="lightpink", lty="blank", ylim=c(0,2.5) )
```


MODEL 2
```{r}
X2=cbind(1, covid$ldensity)
n2=nrow(X2)
k2=ncol(X2)
y2=matrix(covid$lncases,n2,1)

z2=Linv%*%y2
w2=Linv%*%X2
#qr decomposition
qr2=qr(w2)

#check for posterior propriety
stopifnot(n2>k2, qr2$rank==k2)

#calculate posterior hyperparameters
Rinv2=solve(qr.R(qr2))#qr.r extracts the upper triangular matrix
vbeta2=Rinv2%*%t(Rinv2)
betahat2=qr.solve(qr2,z2)
df2=n2-k2
e2=qr.resid(qr2,z2)
sig2=sum(e2^2)/df2

#simulate from posterior
niter=1000
sigma2=sqrt(1/rgamma(niter,df2/2,df2*sig2/2))
beta2=matrix(betahat2,niter,k2, byrow=T)+sigma2*mvrnorm(niter, rep(0,k2),vbeta2)

colMeans(beta2)
betahat2
```

```{r}
#plot parameter estimates m2
par(las=2)
par(mfrow=c(1,3))
hist(sigma2,100, freq=F, main="Standard deviation", xlab=expression(sigma), col="lightblue", lty="blank",ylim=c(0,0.2))
hist(beta2[,1],100, freq=F, main="Intercept", xlab=expression(mu), col="lightpink", lty="blank", ylim=c(0,0.6))
hist(beta2[,2],100, freq=F, main="Beta posterior draws", xlab=expression(beta), col="lightgreen", lty="blank", cex.axis=0.9,ylim=c(0,5))
```



MODEL 3


```{r}
eta5=model.matrix(~-1+as.factor(Group),covid)
eta5[eta5[,5]==1,]=rep(-1,5)
eta5=cbind(1,eta5[,1:4])
X3=eta5
#X3 = cbind(1, covid$Group)
n3=nrow(X3)
k3=ncol(X3)
y3=matrix(covid$lncases,n3,1)

z3=Linv%*%y3
w3=Linv%*%X3
#qr decomposition
qr3=qr(w3)

#check for posterior propriety
stopifnot(n3>k3, qr3$rank==k3)

#calculate posterior hyperparameters
Rinv3=solve(qr.R(qr3))#qr.r extracts the upper triangular matrix
vbeta3=Rinv3%*%t(Rinv3)
betahat3=qr.solve(qr3,z3)
df3=n3-k3
e3=qr.resid(qr3,z3)
sig3=sum(e3^2)/df3

#simulate from posterior
niter=1000
sigma3=sqrt(1/rgamma(niter,df3/2,df3*sig3/2))
beta3=matrix(betahat3,niter,k3, byrow=T)+sigma3*mvrnorm(niter, rep(0,k3),vbeta3)
#apply(beta3,2,quantile,c(0.05,0.5,0.95))
colMeans(beta3)
betahat3
```

```{r}
#plot parameter estimates m3
par(las=2,mfrow=c(1,2))
hist(sigma3,100, freq=F, main="Standard deviation", xlab=expression(sigma), col="lightblue", lty="blank")
hist(beta3[,1],100, freq=F, main="Intercept", xlab=expression(mu), col="lightpink", lty="blank" )

par(mfrow=c(1,5),las=2,mar=c(4,2,1.2,0))
hist(beta3[,2],100, freq=F, main="Eta 1 draws", xlab=expression(eta_1), col="red2", lty="blank", cex.axis=0.9)
hist(beta3[,3],100, freq=F, main="Eta 2 draws", xlab=expression(eta_2), col="goldenrod2", lty="blank", cex.axis=0.9)
hist(beta3[,4],100, freq=F, main="Eta 3 draws", xlab=expression(eta_3), col="palegreen3", lty="blank", cex.axis=0.9)
hist(beta3[,5],100, freq=F, main="Eta 4 draws", xlab=expression(eta_4), col="cornflowerblue", lty="blank", cex.axis=0.9)
hist(-(beta3[,2]+beta3[,3]+beta3[,4]+beta3[,5]),100, freq=F, main="Eta 5 draws", xlab=expression(eta_5), col="darkorchid1", lty="blank", cex.axis=0.9)
```


MODEL 4

```{r}
X4=cbind(X3,covid$ldensity)
n4=nrow(X4)
k4=ncol(X4)
y4=matrix(covid$lncases,n4,1)

z4=Linv%*%y4
w4=Linv%*%X4
#qr decomposition
qr4=qr(w4)

#check for posterior propriety
stopifnot(n4>k4, qr4$rank==k4)

#calculate posterior hyperparameters
Rinv4=solve(qr.R(qr4))#qr.r extracts the upper triangular matrix
vbeta4=Rinv4%*%t(Rinv4)
betahat4=qr.solve(qr4,z4)
df4=n4-k4
e4=qr.resid(qr4,z4)
sig4=sum(e4^2)/df4

#simulate from posterior
niter=1000
sigma4=sqrt(1/rgamma(niter,df4/2,df4*sig4/2))
beta4=matrix(betahat4,niter,k4, byrow=T)+sigma4*mvrnorm(niter, rep(0,k4),vbeta4)


betahat4
colMeans(beta4)
```

```{r}
par(las=2)
par(mfrow=c(1,3))
hist(sigma4,100, freq=F, main="Standard deviation", xlab=expression(sigma), col="lightblue", lty="blank")
hist(beta4[,1],100, freq=F, main="Intercept", xlab=expression(mu), col="lightpink", lty="blank")
hist(beta4[,6],100, freq=F, main="Beta draws", xlab=expression(beta), col="lightgreen", lty="blank", cex.axis=0.9)

par(mfrow=c(1,5),mar=c(4,2,1.2,0),las=2)
hist(beta4[,2],100, freq=F, main="Eta 1 draws", xlab=expression(eta_1), col="red2", lty="blank", cex.axis=0.9)
hist(beta4[,3],100, freq=F, main="Eta 2 draws", xlab=expression(eta_2), col="goldenrod2", lty="blank", cex.axis=0.9)
hist(beta4[,4],100, freq=F, main="Eta 3 draws", xlab=expression(eta_3), col="palegreen3", lty="blank", cex.axis=0.9)
hist(beta4[,5],100, freq=F, main="Eta 4 draws", xlab=expression(eta_4), col="cornflowerblue", lty="blank", cex.axis=0.9)
hist(-(beta4[,2]+beta4[,3]+beta4[,4]+beta4[,5]),100, freq=F, main="Eta 5 draws", xlab=expression(eta_5), col="darkorchid1", lty="blank", cex.axis=0.9)
```



################################################################################################
FUNCTIONS 

```{r}
#Predict using model 1
model1=function(i,j,it){
    yij=beta1[it,1]+rnorm(1,0,sigma1*sqrt(1000/covid$Population[i]))
  return(yij)}

SAC_m1=model1(9,1)
SCL_m1=model1(4,2)
VEN_m1=model1(13,3)
SJ_m1=model1(14,4)
LA_m1=model1(1,5)
m1logpreds=round(c(SAC_m1,SCL_m1,VEN_m1,SJ_m1,LA_m1),2)
m1preds=as.integer(exp(c(SAC_m1,SCL_m1,VEN_m1,SJ_m1,LA_m1)))
actual=c(covid$Total.cases[9],covid$Total.cases[4],covid$Total.cases[13],covid$Total.cases[14],covid$Total.cases[1])
actuallog=round(c(covid$lncases[9],covid$lncases[4],covid$lncases[13],covid$lncases[14],covid$lncases[1]),2)

tm1=as.data.frame(cbind(c("SAC", "SCL","VEN", "SJ","LA"),actuallog,m1logpreds,actual,m1preds))
tm1 %>% kable(col.names = c("County","Log total cases", "M1 prediction (log)","Total cases", "M1 prediction"))%>%kable_styling(position="center",bootstrap_options = c("striped"), latex_options = "hold_position")
```

```{r}
#Predict using model 2
model2=function(i,j,it){
    yij=beta2[it,1]+ beta2[it,2]*covid$ldensity[i]+rnorm(1,0,sigma2*sqrt(1000/covid$Population[i]))
  return(yij)}

SAC_m2=model2(9,1)
SCL_m2=model2(4,2)
VEN_m2=model2(13,3)
SJ_m2=model2(14,4)
LA_m2=model2(1,5)
m2logpreds=round(c(SAC_m2,SCL_m2,VEN_m2,SJ_m2,LA_m2),2)
m2preds=as.integer(exp(c(SAC_m2,SCL_m2,VEN_m2,SJ_m2,LA_m2)))

tm2=as.data.frame(cbind(c("SAC", "SCL","VEN", "SJ","LA"),actuallog,m2logpreds,actual,m2preds))
tm2 %>% kable(col.names = c("County","Log total cases", "M2 prediction (log)","Total cases", "M2 prediction"))%>% kable_styling(position="center",bootstrap_options = c("striped"), latex_options = "hold_position")
```

```{r}
#Predict using model 3
model3=function(i,j,it){
  if(j==1){
    eta1=1
    eta2=0
    eta3=0
    eta4=0}
    if(j==2){
    eta1=0
    eta2=1
    eta3=0
    eta4=0}
    if(j==3){
    eta1=0
    eta2=0
    eta3=1
    eta4=0}  
  if(j==4){
    eta1=0
    eta2=0
    eta3=0
    eta4=1}
    if(j==5){
    eta1=-1
    eta2=-1
    eta3=-1
    eta4=-1}
    yij=beta3[it,1]+ beta3[it,2]*eta1+ beta3[it,3]*eta2+beta3[it,4]*eta3+beta3[it,5]*eta4+rnorm(1,0,sigma3*sqrt(1000/covid$Population[i]))
  return(yij)}

SAC_m3=model3(9,1)
SCL_m3=model3(4,2)
VEN_m3=model3(13,3)
SJ_m3=model3(14,4)
LA_m3=model3(1,5)
m3logpreds=round(c(SAC_m3,SCL_m3,VEN_m3,SJ_m3,LA_m3),2)
m3preds=as.integer(exp(c(SAC_m3,SCL_m3,VEN_m3,SJ_m3,LA_m3)))

tm3=as.data.frame(cbind(c("SAC", "SCL","VEN", "SJ","LA"),actuallog,m3logpreds,actual,m3preds))
tm3 %>% kable(col.names = c("County","Log total cases", "M3 prediction (log)","Total cases", "M3 prediction"))%>% kable_styling(position="center",bootstrap_options = c("striped"), latex_options = "hold_position")
```

```{r}
#Predict using model 4
model4=function(i,j,it){
  if(j==1){
    eta1=1
    eta2=0
    eta3=0
    eta4=0}
    if(j==2){
    eta1=0
    eta2=1
    eta3=0
    eta4=0}
    if(j==3){
    eta1=0
    eta2=0
    eta3=1
    eta4=0}  
  if(j==4){
    eta1=0
    eta2=0
    eta3=0
    eta4=1}
    if(j==5){
    eta1=-1
    eta2=-1
    eta3=-1
    eta4=-1}
    yij=beta4[it,1]+beta4[it,2]*eta1+beta4[it,3]*eta2+beta4[it,4]*eta3+beta4[it,5]*eta4+ beta4[it,6]*covid$ldensity[i]+rnorm(1,0,sigma4[it]*sqrt(1000/covid$Population[i]))
  return(yij)}
beta4
SAC_m4=model4(9,1)
SCL_m4=model4(4,2)
VEN_m4=model4(13,3)
SJ_m4=model4(14,4)
LA_m4=model4(1,5)
m4logpreds=round(c(SAC_m4,SCL_m4,VEN_m4,SJ_m4,LA_m4),2)
m4preds=as.integer(exp(c(SAC_m4,SCL_m4,VEN_m4,SJ_m4,LA_m4)))

tm4=as.data.frame(cbind(c("SAC", "SCL","VEN", "SJ","LA"),actuallog,m4logpreds,actual,m4preds))
tm4 %>% kable(col.names = c("County","Log total cases", "M4 prediction (log)","Total cases", "M4 prediction"))%>% kable_styling(position="center",bootstrap_options = c("striped"), latex_options = "hold_position")
```

######################################################################################

SIMULATIONS

```{r}
niter=1000
M1PREDS=matrix(NA, 58, niter)
M2PREDS=matrix(NA, 58, niter)
M3PREDS=matrix(NA, 58, niter)
M4PREDS=matrix(NA, 58, niter)
for(i in 1:niter){
  for(j in 1:58){
  M1PREDS[j,i]=model1(j, covid$Group[j],i)
  M2PREDS[j,i]=model2(j, covid$Group[j],i)
  M3PREDS[j,i]=model3(j, covid$Group[j],i)
  M4PREDS[j,i]=model4(j, covid$Group[j],i)}}
```

```{r}
#plot histograms for six counties using the four models
par(mfrow=c(1,4))
hist(M1PREDS[9,], main="", lty="blank", col="salmon", xlab="M1 predictions")
abline(v=covid$lncases[9], col="red", lwd=1.5)
hist(M2PREDS[9,], lty="blank",main="", col="salmon", xlab="M2 predictions")
abline(v=covid$lncases[9], col="red", lwd=1.5)
hist(M3PREDS[9,], lty="blank",main="", col="salmon",xlab="M3 predictions")
abline(v=covid$lncases[9], col="red", lwd=1.5)
hist(M4PREDS[9,], lty="blank",main="", col="salmon",xlab="M24 predictions")
abline(v=covid$lncases[9], col="red", lwd=1.5)
mtext("Sacramento (Group 1) predictions", side = 1, line = -33, outer = TRUE)

hist(M1PREDS[4,], main="", lty="blank", col="goldenrod", xlab="M1 predictions")
abline(v=covid$lncases[4], col="red", lwd=1.5)
hist(M2PREDS[4,], lty="blank", main="",col="goldenrod",xlab="M2 predictions")
abline(v=covid$lncases[4], col="red", lwd=1.5)
hist(M3PREDS[4,], lty="blank",main="", col="goldenrod",xlab="M3 predictions")
abline(v=covid$lncases[4], col="red", lwd=1.5)
hist(M4PREDS[4,], lty="blank", main="",col="goldenrod",xlab="M4 predictions")
abline(v=covid$lncases[4], col="red", lwd=1.5)
mtext("Santa Clara (Group 2) predictions", side = 1, line = -33, outer = TRUE)

hist(M1PREDS[13,], main="", lty="blank", col="palegreen2", xlab="M1 predictions")
abline(v=covid$lncases[13], col="red", lwd=1.5)
hist(M2PREDS[13,], lty="blank", main="",col="palegreen2",xlab="M2 predictions")
abline(v=covid$lncases[13], col="red", lwd=1.5)
hist(M3PREDS[13,], lty="blank",main="", col="palegreen2",xlab="M3 predictions")
abline(v=covid$lncases[13], col="red", lwd=1.5)
hist(M4PREDS[13,], lty="blank", main="",col="palegreen2",xlab="M4 predictions")
abline(v=covid$lncases[13], col="red", lwd=1.5)
mtext("Ventura (Group 3) predictions", side = 1, line = -33, outer = TRUE)

hist(M1PREDS[14,], main="", lty="blank", col="cornflowerblue", xlab="M1 predictions")
abline(v=covid$lncases[14], col="red", lwd=1.5)
hist(M2PREDS[14,], lty="blank", main="",col="cornflowerblue",xlab="M2 predictions")
abline(v=covid$lncases[14], col="red", lwd=1.5)
hist(M3PREDS[14,], lty="blank",main="", col="cornflowerblue",xlab="M3 predictions")
abline(v=covid$lncases[14], col="red", lwd=1.5)
hist(M4PREDS[14,], lty="blank",main="", col="cornflowerblue",xlab="M4 predictions")
abline(v=covid$lncases[14], col="red", lwd=1.5)
mtext("San Jose (Group 4) predictions", side = 1, line = -34, outer = TRUE)

hist(M1PREDS[1,],  xlim=c(5.5,9.5),main="", lty="blank", col="violet", xlab="M1 predictions")
abline(v=covid$lncases[1], col="red", lwd=1.5)
hist(M2PREDS[1,], main="",lty="blank", col="violet",xlab="M2 predictions")
abline(v=covid$lncases[1], col="red", lwd=1.5)
hist(M3PREDS[1,],  main="",lty="blank", col="violet",xlab="M3 predictions")
abline(v=covid$lncases[1], col="red", lwd=1.5)
hist(M4PREDS[1,], main="",lty="blank", col="violet",xlab="M4 predictions")
abline(v=covid$lncases[1], col="red", lwd=1.5)
mtext("Los Angeles (Group 5) predictions", side = 1, line = -33, outer = TRUE)

hist(M1PREDS[2,], main="", lty="blank", col="violet", xlab="M1 predictions")
abline(v=covid$lncases[2], col="red", lwd=1.5)
hist(M2PREDS[2,],  main="",lty="blank", col="violet",xlab="M2 predictions")
abline(v=covid$lncases[2], col="red", lwd=1.5)
hist(M3PREDS[2,], main="",lty="blank", col="violet",xlab="M3 predictions")
abline(v=covid$lncases[2], col="red", lwd=1.5)
hist(M4PREDS[2,], main="",lty="blank", col="violet",xlab="M4 predictions")
abline(v=covid$lncases[2], col="red", lwd=1.5)
mtext("San Diego (Group 5) predictions", side = 1, line = -33, outer = TRUE)
```


###############################################################################################
MODEL COMPARISON


4. Compare the different models using justifiable criteria. Can you use g-priors?

R-SQUARED VALUES

```{r}
#MODEL 1
Y_1_hat=w1 %*% betahat1 # Compute Y estimates
res1=z1-Y_1_hat # Compute residuals
SSE1=t(res1) %*% res1 #119907.8
#s1=z1-mean(z1)
s1 = z1 - Y_1_hat
SST1=t(s1) %*% s1

sigmahat1=SSE1/(n1-k1) #2103.646
r2_1=1-SSE1/SST1 #0

#MODEL 2
Y_2_hat=w2 %*% betahat2 # Compute Y estimates
res2=z2-Y_2_hat # Compute residuals
SSE2=t(res2) %*% res2 #
s2=z2-Y_2_hat
#SST2=t(s2) %*% s2  #
sigmahat2=SSE2/(n2-k2) #
r2_2=1-SSE2/SST1 #
#MODEL 3
Y_3_hat=w3 %*% betahat3 # Compute Y estimates
res3=z3-Y_3_hat # Compute residuals
SSE3=t(res3) %*% res3 #
s3=z3-Y_3_hat
SST3=t(s3) %*% s3  #
sigmahat3=SSE3/(n3-k3) #
r2_3=1-SSE3/SST1 # 
#MODEL 4
Y_4_hat=w4 %*% betahat4 # Compute Y estimates
res4=z4-Y_4_hat # Compute residuals
SSE4=t(res4) %*% res4 #
s4=z4-Y_4_hat
SST4=t(s4) %*% s4
sigmahat4=SSE4/(n4-k4) #
r2_4=1-SSE4/SST1 #

round(c(r2_1 ,r2_2,r2_3,r2_4),3)
```

AIC BIC 
```{r}
#https://statswithr.github.io/book/bayesian-model-choice.html
bic1=n1*log(1-r2_1)+(k1+1)*log(n1)
bic2=n2*log(1-r2_2)+(k2+1)*log(n2)
bic3=n3*log(1-r2_3)+(k3+1)*log(n3)
bic4=n4*log(1-r2_4)+(k4+1)*log(n4)
round(c(bic1,bic2,bic3,bic4),1)

aic1=n1*log(1-r2_1)+(k1+1)*2
aic2=n2*log(1-r2_2)+(k2+1)*2
aic3=n3*log(1-r2_3)+(k3+1)*2
aic4=n4*log(1-r2_4)+(k4+1)*2
round(c(aic1,aic2,aic3,aic4),1)
```




```{r}
#deviance function, log likelihood (density) 
devi=function(beta,sigm,w){
  -2*sum(dnorm(z1,w%*%beta,sigm, log=TRUE))
}

big_bar_d1=mean(sapply(1:niter,function(it)devi(beta1[it,],sigma1[it],w1))) #bar_d
small_bar_d1=devi(colMeans(beta1),sqrt(mean(sigma1^2)),w1) #d_bar
#pd1=big_bar_d1-small_bar_d1
#DIC1=pd1+big_bar_d1

#version 2, no neg
d1=sapply(1:niter,function(it)devi(beta1[it,],sigma1[it],w1))
pv1=var(d1)/2
DIC1=pv1+big_bar_d1

big_bar_d2=mean(sapply(1:niter,function(it)devi(beta2[it,],sigma2[it],w2))) #bar_d
small_bar_d2=devi(colMeans(beta2),sqrt(mean(sigma2^2)),w2) #d_bar

d2=sapply(1:niter,function(it)devi(beta2[it,],sigma2[it],w2))
pv2=var(d2)/2
DIC2=pv2+big_bar_d2

big_bar_d3=mean(sapply(1:niter,function(it)devi(beta3[it,],sigma3[it],w3))) #bar_d
small_bar_d3=devi(colMeans(beta3),sqrt(mean(sigma3^2)),w3) #d_bar

d3=sapply(1:niter,function(it)devi(beta3[it,],sigma3[it],w3))
pv3=var(d3)/2
DIC3=pv3+big_bar_d3

big_bar_d4=mean(sapply(1:niter,function(it)devi(beta4[it,],sigma4[it],w4))) #bar_d
small_bar_d4=devi(colMeans(beta4),sqrt(mean(sigma4^2)),w1) #d_bar

d4=sapply(1:niter,function(it)devi(beta4[it,],sigma4[it],w4))
pv4=var(d4)/2
DIC4=pv2+big_bar_d4

round(c(DIC1,DIC2,DIC3,DIC4),1)
```


F-TEST

```{r}
#F-test comparing models 2 vs 4 and 3 vs 4
threshold=qf(.95, k4-k2,n4-k4) #2.549763
f.test.stat=(SSE2-SSE4)/
  (sigmahat4*(k4-k2)) #16.26948
c(threshold,f.test.stat)
f.test.stat > threshold
threshold=qf(.95, k4-k3,n4-k4) #4.026631
f.test.stat=(SSE3-SSE4)/
  (sigmahat4*(k4-k3))#44.19527
c(threshold,f.test.stat)
f.test.stat > threshold
```

5. Based on the results from your analysis, are there other models, involving the proposed grouping and the population density, worth considering?

LASSO
```{r}
iter=1000
llars=lars(w4, z4, type="lasso")
cv=lars::cv.lars( w4, z4 ,plot.it=TRUE, se=TRUE)
gnet=cv.glmnet(w4, z4, standardize=TRUE, alpha=1, nfold=10 )
gnet #min lambda estimate from output

# define the burn-in period and initial values 
burnin = 500
initial.beta=rep(-50, dim(w4)[2]) # assigning an extreme initial value for all betas
initial.lambda2=gnet$lambda.min # assigning an extreme initial value for lambda (penalty parameter)
initial.variance=500 # assigning an extreme initial value for variance parameter

# starting the Gibbs sampler here
lasso=blasso(X = w4, # covariate matrix with dimensions 442 x 64
                y = z4,  # response vector with length of 442
                T = niter, # number of iterations
                beta = initial.beta, 
                lambda2 = initial.lambda2,  
                s2 = initial.variance)

# collecting draws for some of the parameters for visualization
coef.lasso=as.data.frame(cbind(iter = seq(iter), 
                              beta1 = lasso$beta[, "b.1"], beta2 = lasso$beta[, "b.2"], 
                              beta3 = lasso$beta[, "b.3"],beta4 = lasso$beta[, "b.4"], 
                              beta5 = lasso$beta[, "b.5"],beta6 = lasso$beta[, "b.6"],
                              variance = lasso$s2, stdev=sqrt(lasso$s2),
                              lambda.square = lasso$lambda2))
colMedians(coef.lasso[-seq(burnin), -1])
#colMedians(lasso$beta[-seq(burnin), -1])
#sum(colMedians(lasso$beta[-seq(burnin), ]) == 0)
```

```{r}
#Predict using the non zero coefficients from lasso
model5=function(i,j,it){
  if(j==1){
    eta1=1
    eta2=0
    eta3=0
    eta4=0}
    if(j==2){
    eta1=0
    eta2=1
    eta3=0
    eta4=0}
    if(j==3){
    eta1=0
    eta2=0
    eta3=1
    eta4=0}  
    if(j==4){
    eta1=0
    eta2=0
    eta3=0
    eta4=1}
    if(j==5){
    eta1=-1
    eta2=-1
    eta3=-1
    eta4=-1}
    yij=coef.lasso[-seq(burnin), -1][it,1]+ coef.lasso[-seq(burnin), -1][it,2]*eta1+ 0*eta2+ 0*eta3+0*eta4+ 
     coef.lasso[-seq(burnin), -1][it,6]*covid$ldensity[i]+
      rnorm(1,0,sqrt(coef.lasso[it,"variance"]*1000/covid$Population[i]))
  return(yij)}

SAC_m5=model5(9,1,niter)
SCL_m5=model5(4,2,niter)
VEN_m5=model5(13,3,niter)
SJ_m5=model5(14,4,niter)
LA_m5=model5(1,5,niter)
m5logpreds=round(c(SAC_m5,SCL_m5,VEN_m5,SJ_m5,LA_m5),2)
m5preds=as.integer(exp(c(SAC_m5,SCL_m5,VEN_m5,SJ_m5,LA_m5)))

tm5=as.data.frame(cbind(c("SAC", "SCL","VEN", "SJ","LA"),actuallog,m5logpreds,actual,m5preds))
tm5 %>% kable(col.names = c("County","Log total cases", "M5 prediction (log)","Total cases", "M4 prediction"))%>% kable_styling(position="center",bootstrap_options = c("striped"), latex_options = "hold_position")

#Simulations
M5PREDS=matrix(NA, 58, niter)
for(i in 1:niter){
  for(j in 1:58){
  M5PREDS[j,i]=model5(j,covid$Group[j],i)}}
```





```{r}
#LASSO PLOTS
par(mfrow=c(1,5), las=2)
hist(M5PREDS[9,], lty="blank",main="", col="indianred2",xlab="Sacramento")
abline(v=covid$lncases[9], col="red", lwd=1.5)
hist(M5PREDS[4,], lty="blank", main="",col="gold",xlab="Santa Clara")
abline(v=covid$lncases[4], col="red", lwd=1.5)
hist(M5PREDS[13,], lty="blank", main="",col="seagreen2",xlab="Ventura")
abline(v=covid$lncases[13], col="red", lwd=1.5)
hist(M5PREDS[14,], lty="blank",main="", col="lightblue",xlab="San Jose")
abline(v=covid$lncases[14], col="red", lwd=1.5)
hist(M5PREDS[1,], main="",lty="blank", col="mediumpurple",xlab="Los Angeles")
abline(v=covid$lncases[1], col="red", lwd=1.5)
```




G PRIORS AND BAYES FACTOR

####################################################################################


BAYES FACTORS
#A Bayes factor greater than 1 can be interpereted as evidence against the compared-to model (the denominator).
```{r}
n=58

g=0.1
BF21=(1+g)^((n-k2-1)/2)/(1+g*(1-r2_2))^((n-1)/2)#comparing M2 against M1(null) 
BF31=(1+g)^((n-k3-1)/2)/(1+g*(1-r2_3))^((n-1)/2) # comparing M3 against M1(null)   
BF41=(1+g)^((n-k4-1)/2)/(1+g*(1-r2_4))^((n-1)/2) # comparing M4 against M1(null)  
BF23=BF21/BF31# comparing M2 against M3
BF24=BF21/BF41# comparing M2 against M4 
BF34=BF31/BF41# comparing M3 against M4

g0.1.null=c(round(BF21,2), round(BF31,2),round(BF41,2))
g0.1=c(round(BF23,2),round(BF24,2),round(BF34,2))

g=1
BF21=(1+g)^((n-k2-1)/2)/(1+g*(1-r2_2))^((n-1)/2)#comparing M2 against M1(null) 
BF31=(1+g)^((n-k3-1)/2)/(1+g*(1-r2_3))^((n-1)/2) # comparing M3 against M1(null)   
BF41=(1+g)^((n-k4-1)/2)/(1+g*(1-r2_4))^((n-1)/2) # comparing M4 against M1(null)  
BF23=BF21/BF31# comparing M2 against M3
BF24=BF21/BF41# comparing M2 against M4 
BF34=BF31/BF41# comparing M3 against M4

g1.null=c(round(BF21,2), round(BF31,2),round(BF41,2))
g1=c(round(BF23,2),round(BF24,2),round(BF34,2))

g=10
BF21=(1+g)^((n-k2-1)/2)/(1+g*(1-r2_2))^((n-1)/2)#comparing M2 against M1(null) 
BF31=(1+g)^((n-k3-1)/2)/(1+g*(1-r2_3))^((n-1)/2) # comparing M3 against M1(null)   
BF41=(1+g)^((n-k4-1)/2)/(1+g*(1-r2_4))^((n-1)/2) # comparing M4 against M1(null)  
BF23=BF21/BF31# comparing M2 against M3
BF24=BF21/BF41# comparing M2 against M4 
BF34=BF31/BF41# comparing M3 against M4

g10.null=c(round(BF21,2), round(BF31,2),round(BF41,2))
g10=c(round(BF23,2),round(BF24,2),round(BF34,2))


g=100
BF21=((1+g)^((n-k2-1)/2))/(1+g*(1-r2_2))^((n-1)/2)#comparing M2 against M1(null) 
BF31=(1+g)^((n-k3-1)/2)/(1+g*(1-r2_3))^((n-1)/2) # comparing M3 against M1(null)   
BF41=(1+g)^((n-k4-1)/2)/(1+g*(1-r2_4))^((n-1)/2) # comparing M4 against M1(null)  
BF23=BF21/BF31# comparing M2 against M3
BF24=BF21/BF41# comparing M2 against M4 
BF34=BF31/BF41# comparing M3 against M4

g100.null=c(round(BF21,2), round(BF31,2),round(BF41,2))
g100=c(round(BF23,2),round(BF24,2),round(BF34,2))


g=1000
BF21=((1+g)^((n-k2-1)/2))/(1+g*(1-r2_2))^((n-1)/2)#comparing M2 against M1(null) 
BF31=(1+g)^((n-k3-1)/2)/(1+g*(1-r2_3))^((n-1)/2) # comparing M3 against M1(null)   
BF41=(1+g)^((n-k4-1)/2)/(1+g*(1-r2_4))^((n-1)/2) # comparing M4 against M1(null)  
BF23=BF21/BF31# comparing M2 against M3
BF24=BF21/BF41# comparing M2 against M4 
BF34=BF31/BF41# comparing M3 against M4

g1000.null=c(round(BF21,2), round(BF31,2),round(BF41,2))
g1000=c(round(BF23,2),round(BF24,2),round(BF34,2))


g0.1.null
g1.null
g10.null
g100.null
g1000.null
g0.1
g1
g10
g100
g1000

g=10000000000000000
BF21=((1+g)^((n-k2-1)/2))/(1+g*(1-r2_2))^((n-1)/2)#comparing M2 against M1(null) 
BF31=(1+g)^((n-k3-1)/2)/(1+g*(1-r2_3))^((n-1)/2) # comparing M3 against M1(null)   
BF41=(1+g)^((n-k4-1)/2)/(1+g*(1-r2_4))^((n-1)/2) # comparing M4 against M1(null)  
BF23=BF21/BF31# comparing M2 against M3
BF24=BF21/BF41# comparing M2 against M4 
BF34=BF31/BF41# comparing M3 against M4


c(round(BF23,2),round(BF24,2),round(BF34,2))


0.1/(1+0.1)
1/(1+1)
10/(1+10)
100/(100+1)
```
